---
title: "Final Assignment"
author: "Ozan Kaya"
date: "April 6, 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, comment = FALSE, message = FALSE)
library(data.table)
library(ggplot2)
library(stringr)
library(ggpubr)
library(animation)
library(gifski)

```

## Introduction

When we take a look at the yearly summary of the gaming data, we can see that the avg and peak values are extremely skewed. The right tail is rather long. It is not surprising to see this since we should expect some of the games to dominate the gaming world. There should be also many games that have failed to peak the interest of gamers. Therefore, averages misses important information if we are to make an analysis of the data we have. We should focus on the extremes.


```{r}
games <- fread('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-03-16/games.csv')
games_alt <- fread("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-07-30/video_games.csv")

games_merged <- merge(games, games_alt, by.x = "gamename", by.y = "game", all.x = TRUE)
#games_merged <- games_merged[complete.cases(games_merged$release_date)]

#str(games_merged)
# we may need simple transformations to make our life easier.
games_merged$month <- factor(games_merged$month, levels = month.name)
games_merged[, avg_peak_perc := as.numeric(str_remove(avg_peak_perc, "%"))/100]
games_merged$release_year <- str_sub(games_merged$release_date, 
                                     str_count(games_merged$release_date) - 3, str_count(games_merged$release_date) )
games_merged$release_year <- as.numeric(games_merged$release_year)
```

```{r, results = 'asis'}
kable(games_merged[, as.list(summary(avg)), by = year], format = "html", digits = 2, caption = "Summary of Average Gamer Count")  

kable(games_merged[, as.list(summary(peak)), by = year], format = "html", digits = 2, caption = "Summary of Peak Gamer Count")  
```

We can see below that the maximum of average and peak gamer counts resembles each other throughout the years. The obvious peak is on 2017 and 2018 and except these two years, averages seems to show a steady increase. We can now take a look at the monthly figures and then try to figure out the most played games in our dataset.

```{r}
avg_gamers <- ggplot(games_merged[, max(avg)/10^6, by = year], aes(x = year, y = V1)) + geom_col() + 
  ylab("Gamers in millions") + xlab("Year") + labs(title = "Average Gamer Count in Simultaneous Gaming") +
  theme_bw()
max_gamers <- ggplot(games_merged[, max(peak)/10^6, by = year], aes(x = year, y = V1)) + geom_col() + 
  ylab("Gamers in millions") + xlab("Year") + labs(title = "Maximum Gamer Count in Simultaneous Gaming") +
  theme_bw()

ggarrange(avg_gamers, max_gamers, ncol = 2, nrow = 1)


```


Monthly figures shows a similar outcome below. However the monthly variance seems to be rather high. It doesn't seem to follow a strong seasonal pattern but we should investigate this further (see Appendix 2).  I would guess the monthly spikes we see from now and then causes by new releases. But in order to  understand what is going on, we should take a better look at the most played games at these dates.

## Games

```{r, results='asis'}

kable(games_merged[, .SD[which.max(peak)], by = year][order(-year)][, 
            .(gamename, peak, gaming_date, release_date)], format = "html", digits = 2, caption = "Peak Gamer Cout per Year")

kable(games_merged[, .SD[which.max(avg)], by = year][order(-year)][, 
            .(gamename, avg, gaming_date, release_date)], format = "html", digits = 2, caption = "Average Gamer Count per Year")

```

The above information suggests that the top games that drives the peak and average numbers we see are Counter-Strike, Dota 2, PUBG and less so the Cyberpunk and Fallout 4 (from an offline query). Note that Fallout and Cyberpunk are singleplayer games and they tend to have a lower concentration of players on a given time since they don't require cooperation of any kind. Something else that we could point out is that PUBG's success seems to come just after the release of the game and shwos a steady decline in the following months. Dota 2 on the other hand seems to have built its fan base over the years. 

```{r}
monthly_avg <- games_merged[, .SD[which.max(avg)], by = gaming_date][order(-gaming_date)][, 
                                                             .(gamename, avg, gaming_date, release_date)]

monthly_avg_plot <- ggplot(monthly_avg, aes(x = gaming_date, y = avg, fill = gamename)) + geom_col() + theme_bw() + 
  ylab("Gamers in millions") + xlab("Month") + labs(title = "Average Gamer Count in Simultaneous Gaming", 
                                                    fill = "Games") + 
  scale_fill_brewer(type = "qual", palette = "Set1") + theme_bw()

monthly_avg_plot

```

The above graph shows the difference in maximum average gamer count per month from 2012 to 2021. We have seen these values in a yearly level but we didn't checked which games were actually played at a monthly level. Adding color as an extra dimension offers us a nice and easy visualization to this end. 

```{r}
games_merged$release_dummy <- ifelse(games_merged$release_year < 2010, "pre-2010", games_merged$release_year)

peak_combined <- ggplot(games_merged[, .SD[which.max(peak)], 
                                    by = .(year, gamename)][, .(gamename, peak, year, release_dummy)],
                       aes(x = year, y = peak/10^6, color = release_dummy )) + 
  geom_point() + scale_color_brewer(palette = "Paired") + 
  ylab("Gamers in millions") + xlab("Year") + labs(color = "Release Year", 
  title = "Maximum number of gamers per game over the years")

avg_combined <- ggplot(games_merged[, .SD[which.max(avg)], 
                                     by = .(year, gamename)][, .(gamename, avg, year, release_dummy)],
                        aes(x = year, y = avg/10^6, color = release_dummy )) + 
  geom_point() + scale_color_brewer(palette = "Paired") + 
  ylab("Gamers in millions") + xlab("Year") + labs(color = "Release Year", 
                                                   title = "Average number of gamers per game over the years")
ggarrange(avg_combined, peak_combined, ncol = 2, nrow = 1)


```

This is another representation of the playtime PUBG (released at 2017) enjoys over the years. We can also see Dota's (released at 2013) prominence over the years as well as Counter-Strike's (released at 2012) comeback. There are many games that are below the 200.000 average gamer count at the same time and between 100.000 and 200.000, we can see some familiar names such as Fallout 4, Cyberpunk 2077, Civilization 5, GTA 5, Rust and Terraria. 

## Results

I will speculate here that the interest on PUBG was arguably exogenous. Namely, new gamers' seem to have started playing the game on 2017 and 2018 (perhaps changed platforms) and leave the scene after their interest has burned out (Perhaps returned to their otherwise favourite gaming platform). The other most obvious result is the dominance of multiplayer games for our metrics. If we  keep looking at maximum average and peak player count at a given time, we will most likely keep seeing multiplayer or co-op games. If we were to check the total amount of playtime, singleplayer games might had had a shot, especially when they were first released. 


## Appendix 1

So far, I haven't directly pointed out to the relationship between avg and peak values of a game at a given time. But they usually follow a percentage point relationship that is also given to us as a variable. This percentage point seems to be around 0.42 on average throughout our dataset and this translates into a slope of roughly 1/2 if we plot peak on the x axis and avg on the y axis.  As an additional task, I will try to visualize what happens to individual games' avg and peak numbers month over month with also plotting the gain variable to see the differences in the avg variable vis-a-vis the past month.



```{r, include= FALSE}
ani.options(interval = 1)
saveGIF({
  for(i in as.character(sort(unique(games_merged$gaming_date)))) {
    print(ggplot(games_merged[gaming_date == i], aes(peak/10^6, avg/10^6)) + 
      geom_point(aes(color = gain)) + geom_smooth(method = "lm", se = FALSE) + 
      scale_color_distiller(palette = "Spectral") + ylab("Average Number of Players (in mm)") + 
      xlab("Maximum Number of Players (in mm)") +  labs(title = i) + theme_bw())
  }
})
# gganimate fails due to some dependency problems. I opt to produce an animation instead.

```

## Appendix 2

If we take look at the seasonal changes in our data, we can also see that there is a slight seasonality for our games. The major effect is due to the time trend but we also see a very slight increase in peak gamer counts in Counter-Strike around Spring and around Summer in PUBG and around the late Winter in Dota 2. The effects are rather small but it may be still worthwhile to mention them.

```{r}
my_plot.decomposed.ts = function(x, title="", ...) {
  xx <- x$x
  if (is.null(xx)) 
    xx <- with(x, if (type == "additive") 
      random + trend + seasonal
      else random * trend * seasonal)
  plot(cbind(observed = xx, trend = x$trend, seasonal = x$seasonal, random = x$random), 
       main=title, ...)
}

cs_ts <- ts(games_merged[gamename ==  "Counter-Strike: Global Offensive", .(gaming_date,peak)][order(gaming_date)]$peak,
            frequency = 12, start = c(2012,7))
decompsoed_cs <- decompose(cs_ts)

pubg_ts <- ts(games_merged[gamename ==  "PLAYERUNKNOWN'S BATTLEGROUNDS", .(gaming_date,peak)][order(gaming_date)]$peak,
            frequency = 12, start = c(2012,7))
decompsoed_pubg <- decompose(pubg_ts)

dota_ts <- ts(games_merged[gamename ==  "Dota 2", .(gaming_date,peak)][order(gaming_date)]$peak,
            frequency = 12, start = c(2012,7))
decompsoed_dota <- decompose(dota_ts)

```

```{r, out.width='33%'}
my_plot.decomposed.ts(decompsoed_cs, title = "Counter-Strike")
my_plot.decomposed.ts(decompsoed_pubg, title = "PUBG")
my_plot.decomposed.ts(decompsoed_dota, title = "Dota 2")

```